Compilation and Backend-Independent Vectorization for Multi-Party Computation

### 2022-05-02 overleaf/tech_report/introduction.tex
Freezing tech report.

MPC theory dates back to the early 1980-ies~\cite{FOCS:Yao82b,STOC:GolMicWig87,STOC:BenGolWig88,STOC:ChaCreDam88}.
Long the realm of theoretical cryptography, MPC has seen significant advances in programming technology in recent years.
These advances bring MPC closer to practice and wider applicability ---
MPC technology has been employed in real-world scenarios such as auctions~\cite{FC:BCDGJK09}, biometric identification~\cite{ESORICS:BlaGas11},
and privacy-preserving machine learning~\cite{SP:MohZha17,CCS:MohRin18}.
The goal is to bring the technology to a level where programmers can write \emph{secure} and \emph{efficient} programs without
commanding extensive knowledge of cryptographic primitives.

@@

We formalize the MPC Source~\cite{Ishaq:2019} intermediate representation and emphasize optimization over MPC Source. As in classical compilers, we envision different front ends (e.g., our front end IMP Source, other) compiling into MPC Source. MPC Source is particularly suitable for optimizations such as protocol mixing~\cite{Ishaq:2019, Fang:2022} and SIMD-vectorization,
which takes advantage of amortization at the circuit level. The MPC Source IR exposes the \emph{linear structure} of MPC programs, which simplifies program analysis; this is
in contrast to source, which has if-then-else constructs. In the same time, MPC Source is sufficiently ``high-level''  to support analysis and optimizations that take into account
control and data flow in a specific program. MPC Source facilitates cost modeling. 
%MPC Source is small in size and analysis is tractable, as opposed to analysis over an unrolled circuit~\cite{Ishaq:2019}. 
Again as in classical compilers, we envision translation of MPC Source (optimized or unoptimized) into MOTION, SPDZ, or other back-end code.

@@

\subsection{Our Contribution} In this paper, we develop a compiler framework that takes a Python-like routine and produces MOTION code: we describe (a) the IMP Source language, its syntax and semantic restrictions, (b) translation from IMP Source into MPC Source, (c) a specific backend-independent optimization: novel SIMD-vectorization on MPC Source, and (d) translation from MPC Source into MOTION code. %\ana{I think we have to add more on MOTION, specifically mixing and amortization.}

We focus on the MOTION framework as our back-end for several reasons. First, it is the state of the art in terms of performance~\cite{Braun:2022}. Second, it provides an API over efficient implementation for a wide variety of cryptographic operations in three different protocols ---  Arithmetic GMW, Boolean GMW, and BMR --- which allows for protocol mixing~\cite{Ishaq:2019, Fang:2022}, a known backend-independent optimization. Third, MOTION provides API for SIMD-level operations, which amortize cost and lead to significant improvement in memory footprint and
throughput~\cite{NDSS:DemSchZoh15, CCS:ABFKLO18, Braun:2022}. It enables MPC Source-level vectorization, a key focus of this paper.

Our second contribution is an analytical model for cost estimation of amortized schedules. Originally, we hoped that optimal scheduling (under our model, which essentially minimizes the length of the schedule) was tractable, as the problem appeared simpler than the classical scheduling problem. Unfortunately, we show that optimal scheduling is NP-hard via a reduction to the Shortest Common Supersequence (SCS) problem. Cost modeling is important as it drives not only vectorization but optimizations such as protocol mixing and scheduling as well~\cite{Ishaq:2019, Fang:2022}.

Our most important contribution is the implementation and evaluation of the compiler framework. We demonstrate expressivity of the source language by running the compiler on \ana{X} programs with interleaved if- and for-statements; these include classical MPC benchmarks such as PSI and Biometric matching, as well as kMeans, Histogram, and others. \ana{Summarize results.}

@@

>>>>

MPC theory dates back to the early 1980-ies~\cite{FOCS:Yao82b,STOC:GolMicWig87,STOC:BenGolWig88,STOC:ChaCreDam88}.
Long the realm of theoretical cryptography, MPC has seen significant advances in programming technology in recent years.
These advances bring MPC closer to practice and wider applicability ---
MPC technology has been employed in real-world scenarios such as auctions~\cite{FC:BCDGJK09}, biometric identification~\cite{ESORICS:BlaGas11},
and privacy-preserving machine learning~\cite{SP:MohZha17,CCS:MohRin18}.
The goal is to improve technology so that programmers can write \emph{secure} and \emph{efficient} programs without
commanding extensive knowledge of cryptographic primitives.

@@

We formalize the MPC Source~\cite{Ishaq:2019} intermediate representation and emphasize optimization over MPC Source. As in classical compilers, we envision different front ends (e.g., our front end IMP Source, other) compiling into MPC Source. MPC Source is particularly suitable for optimizations such as protocol mixing~\cite{CCS:BDKKS18,Ishaq:2019, Fang:2022} and SIMD-vectorization,
which takes advantage of amortization at the circuit level. The MPC Source IR exposes the \emph{linear structure} of MPC programs, which simplifies program analysis; this is
in contrast to source, which has if-then-else constructs. In the same time, MPC Source is sufficiently ``high-level''  to support analysis and optimizations that take into account
control and data flow in a specific program. MPC Source facilitates cost modeling. 
%MPC Source is small in size and analysis is tractable, as opposed to analysis over an unrolled circuit~\cite{Ishaq:2019}. 
Again as in classical compilers, we envision translation of MPC Source (optimized or unoptimized) into MOTION, SPDZ, or other back-end code.

@@

\subsection{Our Contribution} In this paper, we develop a compiler framework that takes a Python-like routine and produces MOTION code: we describe (a) the IMP Source language, its syntax and semantic restrictions, (b) translation into MPC Source, (c) a specific backend-independent optimization: novel SIMD-vectorization on MPC Source, and (d) translation from MPC Source into MOTION code. %\ana{I think we have to add more on MOTION, specifically mixing and amortization.}

We focus on the MOTION framework as our back-end for several reasons. First, it is the state of the art in terms of performance~\cite{Braun:2022}. Second, it provides an API over efficient implementation for a wide variety of cryptographic operations in three different protocols ---  Arithmetic GMW, Boolean GMW, and BMR --- which allows for protocol mixing~\cite{CCS:BDKKS18,Ishaq:2019, Fang:2022}, a known backend-independent optimization. Third, MOTION provides API for SIMD-level operations, which amortize cost and lead to significant improvement in memory footprint and
throughput~\cite{NDSS:DemSchZoh15, CCS:ABFKLO18, Braun:2022}. It enables MPC Source-level vectorization, a key focus of this paper.

Our second contribution is an analytical model for cost estimation of amortized schedules. Originally, we hoped that optimal scheduling (under our model, which essentially minimizes the length of the schedule) was tractable, as the problem appeared simpler than the classical scheduling problem. Unfortunately, we show that optimal scheduling is NP-hard via a reduction to the Shortest Common Supersequence (SCS) problem. Cost modeling is important as it drives not only vectorization but optimizations such as protocol mixing and scheduling as well~\cite{CCS:BDKKS18,Ishaq:2019, Fang:2022}.

Our most important contribution is the implementation and evaluation of the compiler framework. We demonstrate expressivity of the source language by running the compiler on 16 programs with interleaved if- and for-statements; these include classical MPC benchmarks such as PSI and Biometric matching, as well as kMeans, Histogram, and other examples from the literature. 
Our compiler takes the routine and generated \emph{non-vectorized} MOTION code (from MPC Source on the picture above). It then optimizes MPC Source and generates 
\emph{vectorized} MOTION code (from Optimized MPC Source). We then run the two versions using Boolean GMW and the BMR protocols. (MOTION, which is designed for 
protocol mixing, supports Arithmetic GMW, however, it does not implement Comparison (CMP) and Multiplexing (MUX) as they would be rather inefficient.)
In our experiments vectorized code exhibits 24x improvement on average in circuit generation time for Boolean GMW (20x for BMR), 7x reduction 
in communication (2x), 97x reduction in number of gates (91x), 4x improvement in setup time (23x) and 21x improvement in online time (18x).
%\ana{Summarize results.}

@@

All our code, including benchmark Python-like code, compilation phases, and generated MOTION code is available on Github. We do not provide the link only to preserve anonymity, 
however, we will gladly make it available upon request from reviewers. The Github setup generates graphs and intermediate code for each program along each compiler phase and runs MOTION on 
small inputs to generate tables of data (the experiments in the paper run on real LAN and WAN). We plan to release the link and code, which we believe will be useful to researchers.

### 2022-05-02 overleaf/tech_report/overview.tex
Freezing tech report.

Our compiler takes essentially standard IMP~\ana{Need citation here.}
syntax and imposes certain semantic restrictions. (We detail the restrictions in the following sections.)
The programmer writes an iterative program and annotates certain inputs
and outputs as \emph{shared}. In the example arrays {\sf C} and {\sf S}
are \texttt{shared}, meaning that they store shares, however, the array sizes {\sf D} and
{\sf N} respectively are plaintext. The code iterates over the entries in the
database and computes the Euclidean distance of the current
entry ${\sf S}[i]$ and {\sf C} (its square actually). The program returns the index of the vector that gives
the best match plus the corresponding sum of squares.

@@

We turn to our analytical model to compute the \emph{cost} of the iterative program. Assuming
cost $\beta$ for a local MPC operation (essentially just ADD) and cost $\alpha$ for a remote
MPC operation (e.g., MUX, CMP, and remaining operations), the cost of the iterative schedule
will be $ND(2\alpha+\beta) + N(3\alpha$).

A key contribution is the vectorizing transformation. We can compute all $N*D$
subtraction operations (line 9 in (b)) in a single SIMD instruction; similarly we can compute
all multiplication operations (line 10) in a single SIMD instruction. And while computation
of an individual sum remains sequential, we can compute the $N$ sums in parallel.
Our compiler \emph{automatically detects these opportunities and transforms the program}.
It is standard that MPC researchers write vectorized versions of the Biometric
program by hand; we are the first (to the best of our knowledge) to automatically
transform a naive, iterative MPC program into an unintuitive vectorized one.

@@

Thus, the fixed cost of the vectorized program amounts to
$2\alpha_\mathit{fix}$ + $D \beta_\mathit{fix} + N(3\alpha_\mathit{fix})$. (The variable cost is the same in both the vectorized and non-vectorized programs.)
The first term in the sum corresponds to the vectorized
subtraction and multiplication (lines 9-10 in (c)), the second term corresponds to the for-loop on $j$ (lines 12-16) and the third
one corresponds to the remaining for-loops on $i$ (lines 19-25). 
Clearly, $2\alpha_\mathit{fix} + D\beta_\mathit{fix} + N(3\alpha_\mathit{fix}) << ND(2\alpha_\mathit{fix}+\beta_\mathit{fix}) + N3\alpha_\mathit{fix}$.
Empirically, we observe 10x to 50x improvement over un-vectorized Biometric Matching in circuit generation time and setup time and 5x to 30x improvement in online time.
Additionally, the un-vectorized version runs out of memory for $N=256$, while the vectorized one runs with the standard maximal input size $N=4,096$.
\ana{Edit numbers with final experiments.}

>>>>

Our compiler takes essentially standard IMP~\cite{Nipkow2014}
syntax and imposes certain semantic restrictions. (We detail the restrictions in the following sections.)
The programmer writes an iterative program and annotates certain inputs
and outputs as \emph{shared}. In the example arrays {\sf C} and {\sf S}
are \texttt{shared}, meaning that they store shares, however, the array sizes {\sf D} and
{\sf N} respectively are plaintext. The code iterates over the entries in the
database and computes the Euclidean distance of the current
entry ${\sf S}[i]$ and {\sf C} (its square actually). The program returns the index of the vector that gives
the best match plus the corresponding sum of squares.

@@

We turn to our analytical model to compute the \emph{cost} of the iterative program. Assume
cost $\beta$ for a local MPC operation (e.g., ADD in Arithmetic sharing) and cost $\alpha$ for a remote
MPC operation (e.g., MUX, CMP, etc.). Assuming that ADD is $\alpha$ and SUM, CMP and MUX are $\beta$, 
the MPC Source in Listing~\ref{tab:source_and_MPC_source_and_optimized_MPC_source}(b) gives 
rise to an iterative schedule with cost $ND(2\alpha+\beta) + N(3\alpha$).

A key contribution is the vectorizing transformation. We can compute all $N*D$
subtraction operations (line 9 in (b)) in a single SIMD instruction; similarly we can compute
all multiplication operations (line 10) in a single SIMD instruction. And while computation
of an individual sum remains sequential, we can compute the $N$ sums in parallel.
%Our compiler \emph{automatically detects these opportunities and transforms the program}.
%It is standard that MPC researchers write vectorized versions of the Biometric
%program by hand; we are the first (to the best of our knowledge) to automatically
%transform a naive, iterative MPC program into an unintuitive vectorized one.

@@

Thus, the fixed cost of the vectorized program amounts to
$2\alpha_\mathit{fix}$ + $D \beta_\mathit{fix} + N(3\alpha_\mathit{fix})$. (The variable cost is the same in both the vectorized and non-vectorized programs.)
The first term in the sum corresponds to the vectorized
subtraction and multiplication (lines 9-10 in (c)), the second term corresponds to the for-loop on $j$ (lines 12-16) and the third
one corresponds to the remaining for-loops on $i$ (lines 19-25). 
Clearly, $2\alpha_\mathit{fix} + D\beta_\mathit{fix} + N(3\alpha_\mathit{fix}) << ND(2\alpha_\mathit{fix}+\beta_\mathit{fix}) + N3\alpha_\mathit{fix}$.
Empirically, we observe that $\alpha_\mathit{var} \approx 0$ and orders of magnitude improvement (e.g., 12x in online time in GMW). 
Additionally, the un-vectorized version runs out of memory for $N=256$, while the vectorized one runs with the standard maximal input 
size $N=4,096$.
%\ana{Edit numbers with final experiments.}
